---
title: "R Notebook"
output: html_notebook
---


## R libraries for these functions and snippets

* Note that order in which you import dplyr and plyr makes a difference!

```{r}
library(dplyr)
library(plyr)
library(tidyr)
library(readxl)
library (magrittr)
library(tidyverse)
library(chron)
library(zoo)
library(stringr)
library(janitor)
library(lubridate)
library(anytime)
```

## 1. A helper function to import a csv

* This creates a column with a fixed value in a dataframe and specifies which strings to treat as NA

```{r}
# create csv helper function so can add column

read_csv_filename <-function(filename, year) {
    csv_help <- read.csv(file=filename, na.strings=c("", "n/a", "NULL", "NA"), stringsAsFactors=FALSE)
    csv_help$max_year <- year
    csv_help
}


# run the function as follows

df <- read_csv_filename('C/Amber/Desktop/data.csv', 2018)
```

## 2. A function that checks if dataframes have the same field names and datatypes

* This is useful if you want to append dataframes or maintain consistency between multiple similar dataframes, such as multiple years of the same dataset

```{r}
# check if column names and types are the same across all similar datasets

check_df_fields <- function(dflist) {
    
    # takes a list of dataframes as input
    # uses Janitor compare_df_cols_same to check if colnames and types are the same
    
    # a evaluates to TRUE or FALSE
    a <- compare_df_cols_same(dflist, bind_method = "rbind")
    
    
    # set b as 'same', this updates if not the same
    b <- 'Same'
    if (isFALSE(a)) {
        
    # if  the cols and types are not the same b is updated to a dataframe examining differences
        b <- compare_df_cols(dflist, return = "mismatch", bind_method = "rbind")
        }
    return(b)
        
   }   


# run as follows for list df_list

# list
df_list <- list(df=df, df1=df1, df2=df2)

check_df_fields(df_list)
```

## 3. Build a helper function for basic data cleaning steps

* You can add any additional operations in here


* It requires a dataframe name and the format that date strings are in

```{r}
# clean up

generic_clean_up <- function(df, dateformatstring) {
    
        # convert all column names to lower case
        
        names(df) <-tolower(names(df))
    
         # convert all date columns to date (for dataset with multiple date columns or one)
        # for columns that have 'date' in title
   
    
        if (is.null(ncol(df[,grepl ("^.+date|^+date|date",names(df))])) ){
            df[,grepl ("^.+date|^+date|date",names(df))] <-as.Date(df[,grepl("^.+date|^+date|date",names(df))], format=dateformatstring)


        } else {
            #df[,grepl ("date",names(df))] <-as.Date(df[,grepl("date",names(df))], format="%d/%m/%Y")
            df[,grepl ("^.+date|^+date|date",names(df))] <-lapply(df[,grepl("^.+date|^+date|date",names(df))], as.Date, format=dateformatstring)

        }
    
            return(df)
}
 


# how to run
df <-generic_clean_up(df, "%d/%m/%Y")

# how to run on list of multiple dataframes, df_list

df_list1 <-lapply(df_list, generic_clean_up)
```

## 4. Convert string column of month and year only to first or last day of month

* format MMYYYY to first or last day of month

```{r}
# use zoo package for partial date and to convert to last day of month
# get rid of frac=1 to convert to first day

MMYYYY_date_convert <- function(df, column, first_or_last) {     
    
        a <- first_or_last    
        df[[column]] = ifelse(is.na(df[[column]]), NA,
            if_else(nchar(df[[column]])==5, paste0('0', df[[column]]), 
                   as.character(df[[column]])))
        
        if (a == 'last') {
            df[[column]] <-as.Date(as.yearmon(df[[column]], "%m%Y"), frac = 1)
    }
    
        if (a == 'first') {
                df[[column]] <-as.Date(as.yearmon(df[[column]], "%m%Y"))
        }
    
        return(df)
    }

# how to use

# fill in dataframe name, date_col name in MMYYYY format and 'first' or 'last' depending on whether you want to convert
# to first of last day of month

df <- MMYYYY_date_convert(df, 'date_col', 'last')
```

## 5. Assign dataframe names back after using lapply to operate on a list of them

* After you have used lapply to apply a function to a list of dataframes, you will need to reassign the dataframe names back


* In this example the original list was called df_list and the list you want to assign names back to is called df_list1

```{r}
# assign old dataframe names to new dataframes in the list
for (i in 1:length(df_list)) {
    assign(names(df_list)[[i]], df_list1[[i]])
}
```

## 6. Using if_else to create a categorical column and then convert to factor

* This example creates a column of bmi categories

```{r}
df$bmi_cat =if_else(df$age <19, 'Too young',
            if_else(df$age>=19 & df$bmivalue <18.5, 'Underweight',
            if_else(df$age>=19 & df$bmivalue >=18.5 & df$bmivalue <25, 'Healthy',
            if_else(df$age>=19 & df$bmivalue >=25 & df$bmivalue <30, 'Overweight',       
            if_else(df$age>=19 & df$bmivalue >=30 & df$bmivalue <40, 'Obese', 
            if_else(df$age>=19 & df$bmivalue >=40, 'Morbidly obese',
                   'Unknown'))))) )


df$bmi_cat <- factor(df$bmi_cat, ordered=TRUE,
                    levels =c("Underweight", "Healthy", "Overweight", "Obese", "Morbidly obese", "Unknown", "Too young"))
```

## 7. Creating an age group field based on defined breaks

* This requires an integer column called 'age'

```{r}
agelabels <- c('Under 19', '19-49', '50-64', '65-74', '75-84', '85+')

df$age_group <- cut(df$age, breaks= c(0, 18, 49, 64, 74, 85, 500), labels=agelabels, right=TRUE)    
```

## 8. Drop or keep dataframe columns by name

```{r}
# to drop columns by name

df <- df[!names(df) %in% c("ID", "Age", "Results")]

# to keep columns by name

df <- df[names(df) %in% c("UniqueID", "PostCode", "Occupation")]
```

## 9. Rename multiple dataframe columns

```{r}
# put old names on the left and new names on the right

ccg_rename_vector <- c('Ethnic.Origin'= 'ethnic_orig',
                  'PersonAge'= 'age',
                  'yearofbirth' = 'year_of_birth',
                  'GPpracticecode' = 'practice_code',
                  'GPpracticename' = 'practice_name',
                  'Date.Of.Birth' = 'dob')
                       

# note that here we specify which library's rename function we are using, because dplyr also has one that works differently
df <- plyr:::rename(df, ccg_rename_vector)
```

## 10. Drop duplicate rows

```{r}
# drop duplicates
df <- distinct(df)
```

## 11. Filter dataframe for specific values

* In this case we want to filter for people aged over 18

```{r}
df1 <- df %>% filter (df$age > 18)
```

## 12. Convert integer year to first day of year

```{r}
# convert integer year to first day of year
df$birth_year <- as.Date(paste(df$birth_year, 1, 1, sep = "-"))
```

## 13. Recode a dataframe column based on value in other column

* In this case, we recode the sex to Male, where the person_id = A12

```{r}
df$sex[df$person_id=='A12'] <- 'Male' 
```

## 14. Move a specific column to the start of a dataframe 

* In this case, person_id

```{r}
df <- df %>% select(person_id, everything())
```

## 15. Change dataframe column order

```{r}
col_order <- c('person_id', 'age_group', 'year', 'test_result_1', 'visit_date')

df <- df[, col_order]
```

## 16. Drop multiple rows based on column values

```{r}
df <- df[!df$person_id %in% c( 'C45','D19', 'F187'), ]
```

## 17. Adding up a sum across a row from a specific column to another one

* In this example we are adding up total appointments from January to December AND we are speciying to count NA as 0


* This example will work for consecutive columns

```{r}
df$total_appointments <-rowSums(subset(df, select=totalappointmentsJanuary:totalappointmentsDecember), na.rm=TRUE) * NA ^ (rowSums(!is.na((subset(df, select=totalappointmentsJanuary:totalappointmentsDecember))))==0)
```

## 18. Applying any operation to all of the dataframes in a list

* Note that this returns a list of dataframes, without the original names


* As before, you will need to assign back the orignal names if you want them

```{r}
df_list2 <-lapply(df_list1, function(df) {
    
    #any function here applied to df

        return(df)
})
```

## 19. Get unique list of person ids from multiple dataframes

* This gets a list from each dataframe, combines the rows and drops duplicates

```{r}
# get list of pt_id who are 50+
person1 <- df1 %>% select(person_id)
person2 <- df2 %>% select(person_id)
person3 <- df3 %>% select(person_id)

# create master for adult social care filtering
person_master <- rbind(person1, person2, person3)
person_master <- distinct(person_master)
```

## 20. Assign the tax year to a date

* There are multiple ways to assign a date to a tax year


* This one requires that your date column is in date format


* We use 1st April as start of tax year. This returns a string in format 2017/2018

```{r}
df$date = as.Date(df$date)

# we use yearmon to extract the month and year and count forward 9 months to get the upper year of the tax year
# we paste the upper year of the tax year minus 1, a forward slash, and the upper year of the tax year
df$date <- paste0(as.integer(as.yearmon(df$date)+ 9/12)-1, '/', as.integer(as.yearmon(df$date)+ 9/12))
```

## 21. Drop rows with NA in one column

```{r}
df <- df %>% drop_na(age_cat)
```

## 22. Extract n characters as string from a field

* Can use this like Excel function 'Right' or 'Left'


* Use negative numbers to count from the right, or positive numbers to count from the left


* In this example we extract the last 4 characters from the appointment date field and save in new variable

```{r}
df$appointment_year <- str_sub(df$appointment_date, -4, -1)
```

## 23. Append extra text to column names 

* In this case we are skipping the first column


* We append the suffix gp_ and the highest value of the year variable in this dataframe to each column


* Note that we choose an underscore as a separator

```{r}
names(df)[-1] <- paste(names(df)[-1] , paste0("gp_", max(df$year)), sep = "_")
```

## 24. Count number of NA values across specified columns in each row of the dataframe 

* In this example, we drop rows where count of NA is greater than a certain threshold


* We want all columns including 'year_of_birth'


* MARGIN = 1 means to apply the function across the row

```{r}
#first count na across year of birth cols as most likely to not be blank
df$sums <- apply(df[,grepl ("^.+year_of_birth|^+year_of_birth|year_of_birth",names(df))], MARGIN = 1, function(x) sum(is.na(x)))
                     
# drop rows
df1 <- df[(df$sums < 6),]                     
```

## 25. Sort dataframe columns alphabetically

* Decereasing= FALSE is the default and means sort from A-Z, change to TRUE to sort from Z-A

```{r}
# order columns
df <- df[,order(colnames(df), decreasing=FALSE)]  
```

## 26. Check min and max of each date column in a quick way

* This creates a dataframe with the min and max of each field with date in the title


* Note the use of the summarise_all function to apply summary function to each column


* Note also the use of as.data.frame to store the results in a dataframe

```{r}
# checkfor nonsensical dates

Max <-df[,grepl ("^.+date|^+date|date",names(df))] %>%
    summarise_all(max, na.rm=TRUE) %>%
    as.data.frame
    Max$Measure = 'Max'

Min <-df[,grepl ("^.+date|^+date|date",names(df))] %>%
    summarise_all(min, na.rm=TRUE) %>%
    as.data.frame
    Min$Measure = 'Min'

Sums <- rbind(Min, Max)

Sums %>% select(Measure, everything())
```

## 27. Find the minimum or maximum value (numeric or dates) across columns in a dataframe row

* Note we have used column numbers to refer to which columns we want to look at


* FUN = min, but can put in any function here, max etc


* na.rm=TRUE means we ignore columns with NA


* We can just put 1, instead of writing out MARGIN=1

```{r}
df$lowest_value <- apply(df[, c(8:15,17:18) ], 1, FUN=min, na.rm=TRUE)
```

## 28. Find the nth lowest date or number across certain columns

* Note that we can change to decreasing=TRUE for nth highest

```{r}
# functions for second and third lowest

minn2<-function(x) x[order(x,decreasing=FALSE)[2]]

minn3<-function(x) x[order(x,decreasing=FALSE)[3]]

df$third_lowest_value <- apply(df[, c(8:15,17:18) ], 1, FUN=minn3)
```

## 29 - Find number of days/weeks/months between two dates (in date format)

* Difftime can do seconds, minutes, hours, days, weeks


* To get months or years, divide/ multiply appropriately but remember the word 'weeks' will still be attached


* Here we have also rounded to 2 decimal places

```{r}
# number of weeks between
df$how_many_days<- round((difftime(df$date_1,df$date_2 , units= 'weeks'),digits=2)

# number of months between
df$how_many_days<- round(((difftime(df$date_1,df$date_2 , units= 'weeks')/52)*12),digits=2)
```
